---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

##Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

##Methods
The data for this assignment is located in the same repository as this Rmarkdown file:  
* Dataset = activity.zip  
The variables included in this dataset are:  
1. steps: Number of steps taking in a 5-minute interval (missing values are coded as.NA)  
2. date: The date on which the measurement was taken in YYYY-MM-DD format  
3. interval: Identifier for the 5-minute interval in which measurement was taken  
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

The following procedures were performed during analysis:
1. Read in the data set and added month and day factor variables   
2. Created a histogram of the total number of steps taken each day   
3. Calculated Mean and median number of steps taken each day   
4. Created a time series plot of the average number of steps taken in each time interval   
5. Determined the 5-minute interval that, on average, contains the maximum number of steps     
6. Created code to impute missing data (NA) with the average steps for the given 5 minute interval   
7. Create a histogram of the total number of steps taken each day after missing values are imputed   
8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends   


## Loading and preprocessing the data
```{r}
#start by removing all current objects
rm(list=ls())

library(tidyr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(data.table)
library(lubridate)
library(gridExtra)
library(xtable)

### read in data and clean up data
        #set working directory to folder containing csv file
        setwd("~/Joe's stuff/Stats and data analysis/R code and data/Reproducible Research - Coursera/RepData_PeerAssessment1")

        #import excel file containing IB data
        data<-read.csv(unzip("activity.zip"))
        download.data.date<-Sys.time()
        print(download.data.date)
# code to check structure and NAs in data       
        #str(data)
        #head(data)
        #table(sum(is.na(data)))
        
```
####Process/transform the data into a format suitable for analysis
```{r}
#create factor variables for month and day
data$date<-ymd(data$date)
data$month<-as.factor(month(data$date, label = TRUE))
data$day<-as.factor(day(data$date))
```

## What is mean total number of steps taken per day?
For this part of the assignment NAs were ignored.
```{r}
total_steps_day<-data%>%group_by(date)%>%summarize(total_steps_each_day=sum(steps, na.rm=TRUE))
mean_total_steps<-mean(total_steps_day$total_steps_each_day, na.rm=TRUE)
```
1. Calculate the total number of steps taken per day
```{r}
#plot showing total number of steps by day ignoring NAs 
ggplot(data%>%group_by(month, day)%>%summarize(total_steps=sum(steps, na.omit=TRUE)), aes(x=day, y=total_steps, fill=month))+
        facet_grid(.~month)+
        geom_bar(stat="identity")+
        theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
```{r}
#plot showing total number of steps by day ignoring NAs 
ggplot(total_steps_day, aes(x=total_steps_each_day))+
        geom_histogram(bins = 100)
```
## What is the average daily activity pattern?
3. Calculate and report the mean and median of the total number of steps taken per day
What is the average daily activity pattern?
```{r}
mean_total_steps<-mean(total_steps_day$total_steps_each_day, na.rm = TRUE)
median_total_steps<-median(total_steps_day$total_steps_each_day, na.rm = TRUE)
```
The mean total steps per day was `r round(mean_total_steps)` and the median total steps per day were `r median_total_steps`.

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}

ggplot(data%>%group_by(interval)%>%summarize(mean_steps=mean(steps, na.rm=TRUE)), aes(x=interval, y=mean_steps))+
        geom_line()+
        theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
mean_int_steps<-data%>%group_by(interval)%>%summarize(mean_interval_steps=mean(steps, na.rm=TRUE))%>%arrange(desc(mean_interval_steps))%>%slice(1)
```


## Imputing missing values
Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as.NA). The presence of missing days may introduce bias into some calculations or summaries of the data.
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
table(sum(is.na(data)))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
# determine mean steps per interval then use this value to replace NAs in that interval
# first determine mean steps for each interval
        # did this up above... mean_int_steps
# combine mean steps by interval with data dataframe so all intervals have mean value for that interval
data<-full_join(data, mean_int_steps, by="interval")

# if steps is NA then replace with mean steps that equals the same interval
# this is done by simply identifying NA then replacing with the mean_interval_steps in the same row
data_na.rm<-transform(data, steps=ifelse(is.na(steps),data$mean_interval_steps[is.na(data$steps)], steps))

# check for NAs
table(sum(is.na(data_na.rm)))

# look at the plot of the data to make sure 
plot_with_na<-ggplot(data%>%group_by(month, day)%>%summarize(total_steps=sum(steps, na.rm=TRUE)), aes(x=day, y=total_steps, fill=month))+
        facet_grid(.~month)+
        geom_bar(stat="identity")+
        theme(axis.text.x = element_text(angle = 90, hjust = 1))

plot_no_na<-ggplot(data_na.rm%>%group_by(month, day)%>%summarize(total_steps=sum(steps, na.rm=TRUE)), aes(x=day, y=total_steps, fill=month))+
        facet_grid(.~month)+
        geom_bar(stat="identity")+
        theme(axis.text.x = element_text(angle = 90, hjust = 1))

grid.arrange(plot_with_na,plot_no_na, ncol=1)
```


4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
#plots showing distribution of total number of steps by day from data with and without NAs 
dist_with_na<-ggplot(total_steps_day, aes(x=total_steps_each_day))+
        geom_histogram(bins = 100)+
        labs(title="Mean Total Steps per Day with NAs")

dist_without_na<-ggplot(data_na.rm%>%group_by(date)%>%summarize(total_steps_each_day=sum(steps, na.rm=TRUE)), aes(x=total_steps_each_day))+
        geom_histogram(bins = 100)+
        labs(title="Mean Total Steps per Day without NAs")

grid.arrange(dist_with_na, dist_without_na, ncol=1)
```
```{r}
# determine total steps per day from data set without NAs
total_steps_day_no_NA<-data_na.rm%>%group_by(date)%>%summarize(total_steps_each_day=sum(steps, na.rm=TRUE))

#create a data.frame
#df<-data.frame(mean_total_steps_with_NA=numeric(), median_total_steps_with_NA=numeric(), mean_total_steps_no_NA=numeric(), median_total_steps_no_NA=numeric(), stringsAsFactors = F)

df<-data.frame(ncol=4, nrow=1)

#df = data.frame(matrix(vector(), 0, 4,
   #             dimnames=list(c(), c("mean_total_steps_with_NA", "median_total_steps_with_NA", "mean_total_steps_no_NA", "median_total_steps_no_NA"))),
    #            stringsAsFactors=F)
meantotal_steps_with_NA<-as.numeric(mean(total_steps_day$total_steps_each_day, na.rm = TRUE))
median_total_steps_with_NA<-as.numeric(median(total_steps_day$total_steps_each_day, na.rm = TRUE))
mean_total_steps_no_NA<-as.numeric(mean(total_steps_day_no_NA$total_steps_each_day, na.rm = TRUE))
median_total_steps_no_NA<-as.numeric(median(total_steps_day_no_NA$total_steps_each_day, na.rm = TRUE))

df$mean_total_steps_with_NA<-as.numeric(mean(total_steps_day$total_steps_each_day, na.rm = TRUE))
df$median_total_steps_with_NA<-as.numeric(median(total_steps_day$total_steps_each_day, na.rm = TRUE))
df$mean_total_steps_no_NA<-as.numeric(mean(total_steps_day_no_NA$total_steps_each_day, na.rm = TRUE))
df$median_total_steps_no_NA<-as.numeric(median(total_steps_day_no_NA$total_steps_each_day, na.rm = TRUE))

df<-select(df, -c(ncol,nrow))
df<-gather(df, "statistic", "value", 1:4)
print(df)
#data.frame(c(mean_total_steps_with_NA, mean_total_steps_no_NA, median_total_steps_with_NA, median_total_steps_no_NA))

```

## Are there differences in activity patterns between weekdays and weekends?

Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.
	1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
	2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
Submitting the Assignmentless
```{r}
data_na.rm$week <- ifelse(weekdays(data_na.rm$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
data_na.rm$week<-as.factor(data_na.rm$week)
```
```{r}

ggplot(data_na.rm%>%group_by(week,interval)%>%summarize(mean_steps=mean(steps, na.rm=TRUE)), aes(x=interval, y=mean_steps))+
        facet_grid(week~.)+
        geom_line()+
        theme(axis.text.x = element_text(angle = 90, hjust = 1))
```